## Practical Machine Learning Course Project

### Qi Fang


### 1. Data loading
```{r}
setwd("~/Desktop/JHU/08_practicalMachineLearning")
rm(list = ls())
library(caret)
library(randomForest)
set.seed(1)
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
train <- read.csv(url(trainUrl), na.strings = c("NA", "#DIV/0!", ""))
test <- read.csv(url(testUrl), na.strings = c("NA", "#DIV/0!", ""))
```


### 2. Data partition
#### I use 60% of the "train" data as training set (ptrain) and 40% if the "train" data as validation set (pval).
```{r}
inTrain <- createDataPartition(y = train$classe, p = 0.6, list = FALSE)
ptrain <- train[inTrain, ]
pval <- train[-inTrain, ]
```


### 3. Data cleaning
#### Remove the variables with near zero variance.
```{r}
nzv <- nearZeroVar(ptrain, saveMetrics = TRUE)
ptrain <- ptrain[, nzv$nzv == FALSE]
pval <- pval[, nzv$nzv == FALSE]
```

#### Remove the variables with mostly NA values (>95% of the values are labeled with NA).
```{r}
mostlyNA <- sapply(ptrain, function(x) mean(is.na(x))) > 0.95
ptrain <- ptrain[, mostlyNA == FALSE]
pval <- pval[, mostlyNA == FALSE]
```

#### Remove the first five variables that are not relavent to the prediciton of classe.
```{r}
names(ptrain)[1:5]
ptrain <- ptrain[, -(1:5)]
pval <- pval[, -(1:5)]
```

#### Note that I choose the features to be removed from only ptrain set, and the pval set is not seen, so the out-of-sample error will not be underestimated. Then I remove those features in both ptrain and pval set。


### 4. Model building
#### Here I build models to predict the classe from the predictors. The in-sample-error is estimated from the ptrain set and the out-sample-error is estimated from the pval set. The model with the smallest out-sample-error is selected.

#### First I fit a random forest model to ptrain set, and use 3-fold cross validation tp find the optimal tuning parameters.
```{r}
fitControl <- trainControl(method = "cv", number = 3, verboseIter = FALSE)
fit_rf <- train(classe ~ ., data = ptrain, method = "rf", trControl = fitControl)
fit_rf$finalModel
```
#### The in-sample-error of random forest model is 0.29%

#### Then I fit a gradient boosted model to ptrain set, and use 3-fold cross validation tp find the optimal tuning parameters.
```{r, results = "hide"}
fitControl <- trainControl(method = "cv", number = 3, verboseIter = FALSE)
fit_boost <- train(classe ~ ., data = ptrain, method = "gbm",trControl = fitControl)
```
```{r}
fit_boost
```
#### The in-sample-error of gradient boosted model is 1 - 0.9831011 = 1.69%. Both models show minimal error rate, and the random forest model is even better. Next, I will evaluate the two models on validation set to estimate the out-sample-error and select the best model.

### 5. Model evaluation and selection
#### I fit the models to validation set, and use the confusion matrix to examine the accuracy.
```{r}
## Random forest model
preds_rf <- predict(fit_rf, newdata = pval)
confusionMatrix(pval$classe, preds_rf)
```
```{r}
## Gradient boosted model
preds_boost <- predict(fit_boost, newdata = pval)
confusionMatrix(pval$classe, preds_boost)
```
#### The accuracy of random forest model is 99.77%, and that of gradient boosted model 98.94%. Both model show excellent accuracy on the unseen validation set, and again random forest model is even better. Thus, in the following section I will fit the random forest model to the whole training set ("train") and predict the classes of testing set ("test").


### 6. Predicting classes of testing set using random forest model
#### Fit the random forest model to the whole training set ("train") and predict the classes of testing set ("test").
```{r}
nzv <- nearZeroVar(train)
train <- train[, -nzv]
test <- test[, -nzv]
mostlyNA <- sapply(train, function(x) mean(is.na(x))) > 0.95
train <- train[, mostlyNA == FALSE]
test <- test[, mostlyNA == FALSE]
train <- train[, -(1:5)]
test <- test[, -(1:5)]
fitControl <- trainControl(method = "cv", number = 3, verboseIter = FALSE)
fit <- train(classe ~ ., data = train, method = "rf", trControl = fitControl)
preds <- predict(fit, newdata = test)
preds <- as.character(preds)
```

#### Finally I output the predictions.
```{r}
pml_write_files <- function(x) {
  n <- length(x)
  for(i in 1:n) {
    filename <- paste0("problem_id_", i, ".txt")
    write.table(x[i], file = filename, quote = FALSE, row.names = FALSE, col.names = FALSE)
  }
}
pml_write_files(preds)
```